A statistically consistent approach to chemical tagging

### 2018-12-17 article/ms.tex
add some text on exp2

We initialised the factor loads and component assignments randomly, and
fit this generated data set using the expectation-maximization algorithm
as described in the previous Section. After 58 iterations
(and a serial processing time of 40 seconds) we
reached our convergence threshold: the log probability increased by
less than $10^{-5}$ with each iteration (\todo{Figure~X}). For this experiment we specified
the number of latent factors $\NumLatentFactors$ and the number of clustering components
$\NumComponents$ to include, but below we repeat this experiment where the number
of latent factors and clustering components is not known.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/exp1-ll-iterations.png}
	\caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations. After \todo{X} iterations the improvement in log-likelihood per
    expectation-maximization cycle was less than $10^{-5}$.}
    \label{fig:exp1-ll-iterations.png}
\end{figure}

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, the
Bayesian Information Criterion \citep{bic}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:bic}
\end{equation} 
\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
\begin{equation}
 \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
\end{equation}
\noindent{}where $Q$ is the total number of model parameters in both equations,
and the unknown constants $\omega$ and $\gamma$ have constraints
$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.

@@

\subsection{Experiment~2: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
\label{sec:experiment-galah-clusters}

>>>>

We initialised the factor loads and component assignments randomly, and
fit this generated data set using the expectation-maximization algorithm
as described in the previous Section. After \todo{58} iterations
(and a serial processing time of \todo{40} seconds) we
reached our convergence threshold: the log probability increased by
less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
the number of latent factors $\NumLatentFactors$ and the number of clustering components
$\NumComponents$ to include, but below we repeat this experiment where the number
of latent factors and clustering components is not known.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/exp1-ll-iterations.png}
	\caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations. After \todo{X} iterations the improvement in log-likelihood per
    expectation-maximization cycle was less than $10^{-5}$.}
    \label{fig:exp1-ll-iterations}
\end{figure}

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, and the
Bayesian Information Criterion \citep{bic}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
\end{equation} 
%\noindent{}and the pseudo-Bayesian Information Criterion \citep{pseudo-bic}
%\begin{equation}
% \textrm{pseudo-BIC} = 6(1 + \gamma)\omega{}Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \label{eq:pseudo-bic}
%\end{equation}
%\noindent{}where $Q$ is the total number of model parameters in both equations,
%and the unknown constants $\omega$ and $\gamma$ have constraints
%$\omega \geq 1$ and $\gamma > 0$. Here we adopt $\gamma = 0.1$ and $\omega = 1$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.

@@

\subsection{Experiment~2: Inferring astrophysically realistic latent factors}

Latent factors can only be identified up to orthogonality. That is to say that
if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
then our estimates of those latent factors $\factorloads_\textrm{est}$ may not
be identical to the true values. For example, the ordering of the estimated factors
could be different from the true factors, and the ordering of the dimensionality
in latent space would be correspondingly different. Another possibility is that
the estimated factor loads could be flipped in sign relative to the true factor
loads, and the scores would similarly be flipped. In both of these situations
(re-ordering or flipped signs) the log likelihood given the data and the
estimated factor loads $\factorloads_\textrm{est}$ would be identical to the
log likelihood given the data and the true factor loads $\factorloads_\textrm{true}$
despite the difference in ordering and sign. These examples illustrate a more 
general property that the factor loads and factor scores can be orthogonally 
rotated by any rotation matrix\footnote{A rotation matrix is valid if 
$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
$\factorloads_\textrm{est}$ could therefore appear very different from the true 
values, but they only differ by an orthogonal rotation.

While it is trivial to find the rotation matrix $\vec{R}$ that rotates the estimated
factor loads to the true factor loads, this property of factor analyzers limits
how we can interpret the latent variables of our model. For example, if we seek
to interpret the factor loads as being something analagous to nucleosynthetic
processes, then the estimated factor loads will need to be rotated to something
that looks astrophysically plausible before we could attribute one latent factor
as being a particular nucleosynthetic process.

In this experiment we use detailed chemical abundances of metal-poor stars as
reported by \citet{Barklem:2003}, and we demonstrate the limitations in
interpreting the latent factors as nucleosynthetic processes. The data contains
14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
as well as Sr, Y, Ba, and Eu ($s$/$r$-process). We performed a grid search for
the optimal number of latent factors $J$ and components $K$, starting from
$J = 1$ and $K = 1$, and trialling up until $J = 8$ and $K = 5$. We initialised
the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
until a convergence threshold of $10^{-5}$ was reached. In Figure~\ref{fig:exp2-gridsearch-contours}
we show the negative log likelihood and BIC recorded for the optimised set of
model parameters for each permutation of $J$ and $K$. Hyperparameters of $J = 5$
and $K = 1$ are preferred by BIC for this data set.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2003} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

Using our preferred model with $J = 5$ and $K = 1$, we rotated the estimated
factor loads in order to be astrophysically plausible. Specifically we constructed
a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
all entries were zero except for the following entries:
\begin{itemize}
	\item the entry corresponding to Al in the first latent factor to represent
		  odd-Z element production,
	\item the entries corresponding to Ca and Mg in the second latent factor to
		  represent $\alpha$ particle production,
	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
		  latent factor to represent the Fe-peak group,
	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
		  to represent products of the slow neutron capture process, and
	\item the entries corresponding to Eu in the fifth latent factor to represent
		  products of the rapid nucleosynthesis process.
\end{itemize}
These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
orthonormal. We optimise the $J$ angles needed to make the estimated factor loads
$\factorloads_\textrm{est}$ as close to the astrophysical factor loads $\factorloads_\textrm{A}$.
In practice we take the product of many  Givens rotation matrices \citep{Givens},
which reduces to Euler angle rotation in three or fewer dimensions. In Figure
\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
preferred model, where the shaded regions indicate the projected uncertainty on
the rotation angles -- not including any uncertainty on the factor loads themselves.

The rotated loads are not precisely the same as the target astrophysical loads
in part because the target loads only seek to \emph{identify} and \emph{order}
factors that can be interpreted within an astrophysical context. For example,
the first factor does show positive contributions to Al, like the target load,
but it also shows a slight negative contribution with respect to alpha elements
(Mg, Ca), and a gradual increase among Fe-peak elements up to Ni. Similarly while
Ca and Mg are indeed positive in the second factor, as targeted, there is a
weak negative contribution to Mn abundances, which was not part of the target
loads. The fourth factor shows a positive contributions to Sr, Y, Ba, and to a
lesser extent, Eu, which is representative of slow neutron capture production.
Similarly, the last component shows a very strong positive contribution towards
Eu, and could be interpreted as production from the rapid nucleosynthesis process.

The relatively wide uncertainty regions on these factors is perhaps not
surprising given that only 61 stars were used in this data set. However, this
experiment hopefully demonstrates that the rotated latent factors have some
interpretability, albeit limited. After accounting for the contributions by all
$J = 5$ factors, the specific scatter remaining in each dimension is shown in
Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
as 0.01-0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08-0.16~dex (Al and Ba).

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2003}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-factors}
\end{figure*}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
	\caption{Specific scatter remaining in the \citet{Barklem:2003} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

\subsection{Experiment~3: A subset of gravitationally bound stellar clusters 
			from \Galah\ and elsewhere}
\label{sec:exp3}

After demonstrating that there is some interpretability in the latent factors,
here we seek to experiment whether gravitationally bound stellar clusters can
be reliably identified using the model described in Section~\ref{sec:model}.
